---
title: "James Bond through the ages"
author: "José María Mateos - http://rinzewind.org/blog-en"
output: html_document
---

The new James Bond movie, [Spectre](http://www.imdb.com/title/tt2379713/), will be out in a few days. I thought it would be a very nice exercise to do a quick check on how did previous movies do in terms of overall quality (as assessed by the critics and / or the public) and box office earnings. More to the point, I wanted to do that just by writing a single R script, and scraping data off the web, just for fun.

Specifically, we want to obtain the following data for each movie:

* Title.
* Year.
* Who played James?
* What did the critics say?
* Box office (corrected for inflation).

I am not taking the director into consideration, sorry. I might be ignorant about that specific detail, but before Sam Mendes directed [Skyfall](http://www.imdb.com/title/tt1074638/) I didn't even care about who was in charge of each movie.

This is the list of resources to get data from:

* [Wikipedia](https://en.wikipedia.org) (of course).

Really, that's it. I was planning on doing a lot of the analysis myself, but it turns out Wikipedia already has a page with all the relevant information condensed. So no need to duplicated any efforts.

(Just for the record, my old workflow used a different Wikipedia page, then went to IMDB, then to the Federal Reserve for info on inflation. Turns out IMDB doesn't let you scrape their site, so I was trying to find some workaround when I saw a new and wonderful Wikipedia page. In any case, please note that there are [undocumented](https://stackoverflow.com/questions/1966503/does-imdb-provide-an-api) and [unoficial](http://www.omdbapi.com/) APIs, if you ever need them).

And this is the list of relevant libraries for this project:

* `RCurl`: for reading HTTPS data.
* `XML`: we'll need it to parse Wikipedia tables using its very handy `readHTMLTable` function.
* `ggplot2`: for nice plots.

So, let's start by loading these libraries:

```{r loadlibs}
library(RCurl)
library(XML)
library(ggplot2)
# Configure the plots
theme_set(theme_bw(16))
```

Let's proceed. [This is the relevant Wikipedia page](https://en.wikipedia.org/wiki/List_of_James_Bond_films), which stores the info in two separate tables (EON or non-EON films; these last ones include Peter Seller's [Casino Royale](http://www.imdb.com/title/tt0061452/?ref_=nm_flmg_act_24)). Let's read and merge them.

```{r wikipedia, cache = TRUE}
doc <- getURL("https://en.wikipedia.org/wiki/List_of_James_Bond_films")
tables <- readHTMLTable(doc, stringsAsFactors = FALSE)
# Relevant tables are 1 and 3
t1 <- tables[[1]]
t1s <- tables[[2]]
t2 <- tables[[3]]
t1 <- t1[, c(1, 2, 3, 8, 9)]
t1s <- t1s[, c(2, 4)]
t1 <- merge(t1, t1s)
t2 <- t2[, c(2, 1, 3, 7, 8, 9)]
t2 <- t2[-1, ]
tab_names <- c("Year", "Title", "Actor", "Boffice", "Budget", "RT")
names(t1) <- tab_names
names(t2) <- tab_names
# Let's keep this, just in case
t1$eon <- TRUE
t2$eon <- FALSE
bond <- rbind(t1, t2)
bond$Year <- as.numeric(bond$Year)
bond <- bond[order(bond$Year), ]

# Now let's correct some errors in the parsing, and assign proper numerical values
# where necessary
bond$Title <- gsub(x = bond$Title, pattern = ".*\\!", replacement = "")
bond$Actor <- sapply(bond$Actor, function(x) substr(x, nchar(x) / 2 + 2, nchar(x)))
bond$RT <- gsub(x = bond$RT, pattern = "%.*", replacement = "")
bond$Boffice <- as.numeric(bond$Boffice)
bond$Budget <- as.numeric(bond$Budget)
bond$RT <- as.numeric(bond$RT)
```

Ok, we have the data. Let's do a quick check:

```{r bond_table}
table1 <- aggregate(RT ~ Actor, bond, length)
table1 <- table1[order(-table1$RT), ]
names(table1) <- c("Actor", "Number of movies")
table1
```

```{r plots, fig.width = 12, fig.height = 10}
plt1 <- ggplot(bond) + 
        geom_smooth(aes(x = RT, y = Boffice), 
                    method = "lm",
                    color = "black") +
       geom_point(aes(x = RT, 
                      y = Boffice, 
                      color = Actor), 
                  size = 6) +
        scale_color_manual(values = rainbow(7),
                           name = "Actor playing Bond") +
        ylab("Box office [M$, adjusted for inflation, values of 2005]\n") +
        xlab("\nRottenTomatoes.com score [%]") +
        geom_text(aes(x = RT, y = Boffice, label = Title), angle = -45,
                  size = 4, hjust = 0, vjust = 2, alpha = 0.4) +
       annotate("text", label = "Source: Wikipedia (https://en.wikipedia.org/wiki/List_of_James_Bond_films)",
                x = 40, y = 900, hjust = 0.32, color = "gray")
print(plt1)
```

This is also telling us something about the profitability of the movie business.